Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-18 paper2.qmd
section:issues,model; process:write,figures

---
title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
      non-dropping-particle: van
    orcid: https://orcid.org/0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
      non-dropping-particle: De
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - Causal analysis
  - Directed Acyclic Graphs
  - Bayesian statistical methods
  - Thurstonian model
  - Comparative judgement
  - Probability
  - Statistical modeling
abstract: |
  This study revisits Thurstone's law of comparative judgments (CJ) by addressing two key limitations in traditional approaches. Firstly, it addresses the overreliance on the assumptions of Thurstone's Case V in the statistical analysis of CJ data. Secondly, it addresses the apparent disconnect between CJ's approach to trait measurement and hypothesis testing. We put forward a systematic approach based on causal analysis and Bayesian statistical methods, which results in a model that facilitates a more comprehensive understanding of the factors influencing CJ experiments while offering a robust statistical translation. The new model accommodates unequal dispersions and correlations between stimuli, enhancing the reliability and validity of CJ's trait estimation, thereby ensuring the accurate measurement and interpretation of comparative data. The paper highlights the relevance of this updated framework for modern empirical research, particularly in education and social sciences. This contribution advances current research methodologies, providing a robust foundation for future applications in diverse fields.
key-points:
  - (to do)
date: last-modified
bibliography: references.bib
notebook-links: global
lightbox: true
citation: true
  # type: article-journal
  # container-title: "Psychometrika"
  # doi: ""
  # # url: https://example.com/summarizing-output
execute: 
  cache: true
  # eval: true
  echo: false
  # output: true
  # include: true
  warning: false
  error: false
  message: false
---

@@

<!-- 2. Example of discriminal processes -->
@fig-discriminal_process illustrates hypothetical discriminal processes along a quality trait continuum for two written texts. The figure indicates that the modal discriminal process for Text B is positioned further along the continuum than that of Text A $(S_{B} > S_{A})$, suggesting that Text B exhibits higher quality. Additionally, the figure highlights that Text B has a broader distribution compared to Text A, which arises from its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.

@@

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference illustrates the distribution of the discriminal difference for the hypothetical texts depicted in panel @fig-discriminal_process. The figure indicates that the judge perceives Text B as having significantly higher quality than Text A. This conclusion rests on two key observations: the positive difference between their modal discriminal processes $(S_{B} - S_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$. As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.

@@

This section examines the two critical issues in the CJ literature that serve as the primary focus of this study. The first is the overreliance on Thurstone’s Case V assumptions in the statistical analysis of CJ data. The second is the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.

@@

<!-- 1. What is the problem with the general form -->
Thurstone observed that the general form of the theory, outlined in @sec-thurstone_theory, created a trait scaling problem. The model required estimating more "unknown" parameters than the available pairwise comparisons [@Thurstone_1927b, pp. 267]. To address this issue and facilitate the practical application of the theory, he developed five cases derived from this general form, each case progressively incorporating additional simplifying assumptions into the model.

<!-- 2. Thurstone's Cases and their simplifying assumptions -->
In Case I, Thurstone assumed that pairs of stimuli maintained a constant correlation across all comparisons. In Case II, he allowed multiple judges to make comparisons instead of restricting evaluations to a single judge. In Case III, he introduced the assumption of zero correlation between stimuli. In Case IV, he assumed stimuli exhibited similar dispersions. Finally, in Case V, he replaced this assumption with the condition that stimuli had equal discriminal dispersions. @tbl-thurstone_cases summarizes the assumptions of the general form and the five cases. For an in-depth discussion of these cases and their progression, refer to @Thurstone_1927b and @Bramley_2008 [pp. 248–253].

@@

<!-- 3. The Case V and its simpler statistical representation: the BTL model -->
Despite relying on the most extensive set of simplifying assumptions [@Bramley_2008, pp. 253; @Kelly_et_al_2022, pp. 677], Case V remains the most widely used case in the CJ literature. This popularity stems mainly from its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. The BTL model mirrors the assumptions of Case V, with one key difference: while Case V assumes a Normal distribution for the stimuli's discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp. 254] (see @tbl-thurstone_cases). This substitution has little impact on the model's estimation or interpretation, as the Normal and Logistic distributions share similar statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp. 16].

@@

<!-- 4. But what are limitations of Case V (and the BTL model)? -->
However, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp. 677]. He explicitly warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], acknowledging that some assumptions could prove problematic when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376]. Consequently, given that modern CJ applications frequently involve such traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice: the assumption of equal dispersion and zero correlation between stimuli.

@@

<!-- 1. What does the equal dispersion between stimuli imply? -->
According to the theory, discrepancies in the discriminal dispersions of stimuli shape the distribution of the discriminal difference, exerting a direct influence on the outcome of pairwise comparisons. @fig-dispersion illustrates this idea using a thought experiment in which a researcher can observe the discriminal processes for the texts shown in @fig-discriminal_process. Furthermore, the figure assumes that the discriminal dispersion for Text A remains constant and that the texts are uncorrelated $(\rho=0)$. The figure reveals that an increase in the uncertainty associated with the perception of Text B in comparison to Text A, $(\sigma_{B}-\sigma_{A})$, broadens the distribution of their discriminal difference. This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B > A)$, ultimately influencing the comparison outcome. Additionally, the figure reveals that when the discriminal dispersions of the texts are equal $(\sigma_{B}-\sigma_{A}=0)$, the discriminal difference is more likely to favor Text B over Text A (shaded gray area), compared to situations where their dispersions differ.

<!-- 2. But this is not the scenario we observe -->
In experimental practice, however, this process occurs in reverse. Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between the stimuli and their respective discriminal processes [@Thurstone_1927a, pp. 373]. For instance, when researchers observe a large sample of outcomes favoring Text B over Text A and correctly assume equal dispersions between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts. This scenario is illustrated in @fig-dispersion, where the model's discriminal difference distribution aligns with the "true" distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$. The accuracy of this estimated discriminal difference ensures reliable estimates of the texts’ discriminal processes [(citation needed?){style="color:red;"}. Therefore, it is reasonable to assume that the outcome's ability to reflect the "true" differences between stimuli largely depends on the validity of the model's assumptions, particularly the assumption of equal dispersions.

@@

<!-- 3. But does it hold?  -->
However, Thurstone contended that the assumption of equal dispersions may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376], as these traits and stimuli can introduce judgment discrepancies due to their unique features [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. Indeed, evidence of this violation may already exist in the CJ literature as misfit statistics, which measure judgment discrepancies associated with specific stimuli [@Pollitt_2004, pp. 12; @Goossens_et_al_2018, pp. 20]. For example, labeling texts as "misfits" indicates that comparisons involving these texts result in more judgment discrepancies than those involving other texts [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018]. These discrepancies, in turn, suggest that the discriminal differences for "misfit" texts have broader distributions, indicating that their discriminal processes may also exhibit more variation than that of other texts. A similar reasoning applies to "misfit" judges, whose evaluations deviate substantially from the shared consensus due to the unique characteristics of the stimuli or the judges themselves. Moreover, these "misfit" judges and their deviations can introduce additional statistical and measurement issues, which we discuss in @sec-theory-issue1b.

@@

<!-- 4. So, what happens when the assumption does not hold? -->
Therefore, incorrectly assuming equal dispersions between stimuli is not without harm, as it can cause the BTL model to introduce significant statistical and measurement issues. For instance, the model may overestimate how accurately the outcome reflects the "true" discriminal differences between stimuli. This overestimation can lead researchers to draw spurious conclusions about these differences [@McElreath_2020, pp. 370] and, by extension, of the stimuli underlying discriminal processes. @fig-dispersion illustrates this issue when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the "true" discriminal difference follows a discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$. Moreover, if researchers recognize that misfit statistics highlight critical differences in dispersions, the usual CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] can unintentionally discard valuable information. This exclusion can introduce bias into trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases are often unpredictable, as they depend on which stimuli are excluded from the analysis.

@@

<!-- 1. What does the zero correlation between stimuli imply? -->
Denoted by $\rho$, the correlation measures the dependence of a judge's perception of the trait in one stimulus on his perception of the same trait in another. Like the discriminal dispersions, this correlation shapes the distribution of the discriminal difference and directly influences the outcomes of pairwise comparisons. @fig-correlation illustrates this concept, again assuming the researcher can observe the discriminal processes for the texts shown in @fig-discriminal_process. The figure also considers that the discriminal dispersions for both texts remain constant.

@fig-correlation reveals that as the correlation between the texts increases, the distribution of their discriminal difference becomes narrower. This narrowing affects the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$, thus influencing the comparison outcome. Furthermore, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less likely to favor Text B over Text A (shaded gray area), compared to scenarios when the texts are highly correlated.

@@

<!-- 2. But this is not the scenario we observe -->
However, in experimental practice, researchers typically follow this process in reverse. For example, when they observe a large sample of outcomes favoring Text B over Text A and correctly assume zero correlation between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts. This scenario is illustrated with @fig-correlation when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\rho=0$. The accuracy of this discriminal difference estimation, in turn, ensures reliable estimates for the discriminal process of the texts [(citation needed?){style="color:red;"}.

@@

<!-- 2. But does it hold?  -->
Notably, Thurstone's Case V and the BTL model assume independent discriminal processes across comparisons. Thurstone attributed this independence to the cancellation of potential judges' biases, driven by two opposing and equally weighted effects occurring during the pairwise comparisons [@Thurstone_1927b, pp. 268]. @Andrich_1978 mathematically demonstrated this cancellation using the BTL model under the assumption of discriminal processes with additive biases. However, it is easy to imagine at least two scenarios where the zero correlation assumption almost certainly does not hold: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli and when an additional hierarchical structure is relevant to the stimuli.

<!-- 3. The first scenario where the assumption does not hold -->
In the first scenario, the intricate aspects of multidimensional, complex traits may introduce dependencies between the stimuli due to certain judges' biases that resist cancellation. Research on text quality suggests that when judges evaluate these traits, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. These additional relevant characteristics, which are unlikely to be equally weighted or opposing, can unevenly influence judges' perceptions, creating biases in their judgments and, ultimately, introducing dependencies between stimuli [@vanderLinden_et_al_2017_II, pp. 346]. For example, this could occur when a judge assessing the argumentative quality of a text places more weight on its grammatical accuracy than other judges, ultimately favoring texts with fewer errors but weaker arguments. While direct evidence for this specific scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the idea that the factors influencing pairwise comparisons may not always cancel out.

@@

<!-- 4. The second scenario where the assumption does not hold -->
In the second scenario, the shared context or inherent connections created by additional hierarchical structures may further introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010]. Although the CJ literature acknowledges the presence of such hierarchical structures, the statistical handling of this extra source of dependency between stimuli has been inadequate. For example, when CJ data includes multiple samples of stimuli from the same individuals, researchers often rely on (average) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. However, this approach can introduce additional statistical and measurement issues, which we discuss in @sec-theory-issue2.

<!-- 5. But what are the challenges when this assumption is violated? -->
[In any case, similar to @sec-theory-issue1a, incorrectly assuming zero correlation between stimuli can lead the BTL model to introduce various statistical and measurement issues. For instance, the model could over- or underestimate the accuracy of the outcome in reflecting the "true" discriminal differences between stimuli. This over- underestimation may result in spurious inferences about these differences and, by extension, about the stimuli's discriminal processes [@Hoyle_et_al_2023, pp. 341]. @fig-correlation also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\rho=0$, while the "true" discriminal difference follows any discontinuous line where $\rho \neq 0$. This missaligment can be due to the overlook of additional relevant traits, such as judges' biases, which cause dimensional mismatches in the BTL model, artificially inflating the reliability of the trait [@Hoyle_et_al_2023, pp. 341] or, even worse, introduce bias into the trait's estimates [@Ackerman_1989]. Furthermore, researchers who exclude judges based on misfit statistics can risk discarding valuable information, further biasing the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. Lastly, researchers who fail to account for hierarchical (grouping) structures can reduce the precision of model parameter estimates, which may amplify the overestimation of the trait's reliability [@Hoyle_et_al_2023, pp. 482].]{style="color:red;"}

@@

[In aggregate, researchers' inadequate handling of violations to the assumptions of equal dispersion and zero correlation between stimuli, along with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, can undermine the reliability of the trait estimates and ultimately compromise its validity [@Perron_et_al_2015, pp. 2]. Consequently, adopting a more systematic and integrated approach to examining what happens when judges compare two stimuli could offer several statistical and measurement benefits, including addressing these issues.]{style="color:red;"}

@@

This section presents a theoretical model for CJ that extends Thurstone's theory. The model systematically incorporates all factors involved when judges make pairwise comparisons. Additionally, the section develops the statistical translation of the theoretical model based on assumptions informed by the CJ theory.

<!-- in a stepwise manner -->

@@

![Theoretical model A1$](/images/png/CJ_TM_A1.png){#fig-CJ_TM_A1 width=29%}

@@

![Theoretical model A2$](/images/png/CJ_TM_A2.png){#fig-CJ_TM_A2 width=40%}

@@

![Theoretical model A3$](/images/png/CJ_TM_A3.png){#fig-CJ_TM_A3 width=60%}

@@

>>>>

---
title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
      non-dropping-particle: van
    orcid: https://orcid.org/0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
      non-dropping-particle: De
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - Causal analysis
  - Directed Acyclic Graphs
  - Bayesian statistical methods
  - Thurstonian model
  - Comparative judgement
  - Probability
  - Statistical modeling
abstract: |
  This study revisits Thurstone's law of comparative judgments (CJ) by addressing two key limitations in traditional approaches. Firstly, it addresses the overreliance on the assumptions of Thurstone's Case V in the statistical analysis of CJ data. Secondly, it addresses the apparent disconnect between CJ's approach to trait measurement and hypothesis testing. We put forward a systematic approach based on causal analysis and Bayesian statistical methods, which results in a model that facilitates a more comprehensive understanding of the factors influencing CJ experiments while offering a robust statistical translation. The new model accommodates unequal dispersions and correlations between stimuli, enhancing the reliability and validity of CJ's trait estimation, thereby ensuring the accurate measurement and interpretation of comparative data. The paper highlights the relevance of this updated framework for modern empirical research, particularly in education and social sciences. This contribution advances current research methodologies, providing a robust foundation for future applications in diverse fields.
key-points:
  - (to do)
date: last-modified
bibliography: references.bib
---

@@

<!-- 2. Example of discriminal processes -->
@fig-discriminal_process illustrates hypothetical discriminal processes along a quality trait continuum for two written texts. The figure indicates that the modal discriminal process for Text B is positioned further along the continuum than that of Text A $(T_{B} > T_{A})$, suggesting that Text B exhibits higher quality. Additionally, the figure highlights that Text B has a broader distribution compared to Text A, which arises from its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.

@@

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference illustrates the distribution of the discriminal difference for the hypothetical texts depicted in @fig-discriminal_process. The figure indicates that the judge perceives Text B as having significantly higher quality than Text A. This conclusion rests on two key observations: the positive difference between their modal discriminal processes $(T_{B} - T_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$. As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.

@@

This section examines the two critical issues in the CJ literature that serve as the primary focus of this study. The first is the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. The second is the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.

@@

<!-- 1. What is the problem with the general form -->
Thurstone noted that the general form of the theory, as delineated in @sec-thurstone_theory, gave rise to a trait scaling problem. The model required estimating more "unknown" parameters than the available pairwise comparisons [@Thurstone_1927b, pp. 267]. To address this issue and facilitate the practical implementation of the theory, he developed five cases derived from this general form, each case progressively incorporated additional simplifying assumptions into the model.

<!-- 2. Thurstone's Cases and their simplifying assumptions -->
In Case I, Thurstone postulated that pairs of stimuli would maintain a constant correlation across all comparisons. In Case II, he allowed multiple judges to undertake comparisons instead of confining evaluations to a single judge. In Case III, he posited that there was no correlation between stimuli. In Case IV, he assumed that the stimuli exhibited similar dispersions. Finally, in Case V, he replaced this assumption with the condition that stimuli had equal discriminal dispersions. @tbl-thurstone_cases summarizes the assumptions of the general form and the five cases. For a detailed discussion of these cases and their progression, refer to @Thurstone_1927b and @Bramley_2008 [pp. 248–253].

@@

<!-- 3. The Case V and its simpler statistical representation: the BTL model -->
Despite relying on the most extensive set of simplifying assumptions [@Bramley_2008, pp. 253; @Kelly_et_al_2022, pp. 677], Case V remains the most widely used case in the CJ literature. This popularity stems mainly from its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. The BTL model mirrors the assumptions of Case V, with one notable distinction: whereas Case V assumes a Normal distribution for the stimuli's discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp. 254] (see @tbl-thurstone_cases). This substitution has little impact on the model's estimation or interpretation, as the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp. 16].

@@

<!-- 4. But what are limitations of Case V (and the BTL model)? -->
However, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp. 677]. He explicitly warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270]. Furthermore, he acknowledged that some assumptions could prove problematic when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376]. Consequently, given that modern CJ applications frequently involve such traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice, namely the assumption of equal dispersion and zero correlation between stimuli.

@@

<!-- 1. What does the equal dispersion between stimuli imply? -->
According to the theory, discrepancies in the discriminal dispersions of stimuli shape the distribution of the discriminal difference, exerting a direct influence on the outcome of pairwise comparisons. @fig-dispersion presents a thought experiment to illustrate this idea. In this experiment, a researcher can observe the discriminal processes for the texts depicted in @fig-discriminal_process. Furthermore, the figure assumes that the discriminal dispersion for Text A remains constant and that the texts are uncorrelated $(\rho=0)$. The figure reveals that an increase in the uncertainty associated with the perception of Text B in comparison to Text A, $(\sigma_{B}-\sigma_{A})$, broadens the distribution of their discriminal difference. This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B > A)$, ultimately influencing the comparison outcome. Additionally, the figure reveals that when the discriminal dispersions of the texts are equal $(\sigma_{B}-\sigma_{A}=0)$, the discriminal difference is more likely to favor Text B over Text A (shaded gray area), compared to situations where their dispersions differ.

<!-- 2. But this is not the scenario we observe -->
In experimental practice, however, this process occurs in reverse. Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between the stimuli and their respective discriminal processes [@Thurstone_1927a, pp. 373]. Therefore, it is reasonable to assume that the outcome's ability to reflect the "true" differences between stimuli largely depends on the validity of the model's assumptions, particularly the assumption of equal dispersions. For instance, when researchers observe a sample of outcomes favoring Text B over Text A and correctly assume equal dispersions between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts. This scenario is illustrated in @fig-dispersion, where the model's discriminal difference distribution aligns with the "true" distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$. The accuracy of these discriminal difference ensures reliable estimates for the texts’ discriminal processes [(citation needed?)]{style="color:red;"}.

@@

<!-- 3. But does it hold?  -->
However, Thurstone argued that the assumption of equal dispersions may not be applicable when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376], as these traits and stimuli can introduce judgment discrepancies due to their unique features [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. Indeed, evidence of this violation may already be present in the CJ literature in the form of misfit statistics, which measure judgment discrepancies associated with specific stimuli [@Pollitt_2004, pp. 12; @Goossens_et_al_2018, pp. 20]. For example, labeling texts as "misfits" indicates that comparisons involving these texts result in more judgment discrepancies than those involving other texts [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018]. These discrepancies, in turn, suggest that the discriminal differences for "misfit" texts have broader distributions, indicating that their discriminal processes may also exhibit more variation than that of other texts. A similar line of reasoning applies to the concept of "misfit" judges, whose evaluations deviate substantially from the shared consensus due to the unique characteristics of the stimuli or the judges themselves. These "misfit" judges and their associated deviations can give rise to additional statistical and measurement issues, which we discuss in more detail in @sec-theory-issue1b.

@@

<!-- 4. So, what happens when the assumption does not hold? -->
Nevertheless, model misspecification, in the form of an erroneous assumption of equal dispersions between stimuli, can give rise to significant statistical and measurement issues. For instance, the model may overestimate the degree to which the outcome accurately reflects the "true" discriminal differences between stimuli. This overestimation can result in researchers drawing spurious conclusions about these differences [@McElreath_2020, pp. 370] and, by extension, about the underlying discriminal processes of stimuli. @fig-dispersion illustrates this issue when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the "true" discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.

Additionally, if researchers recognize that misfit statistics highlight these critical differences in dispersions, the conventional CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] can unintentionally discard valuable information. Such exclusions can introduce bias into trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases are often unpredictable, as they depend on which stimuli are excluded from the analysis.

@@

<!-- 1. What does the zero correlation between stimuli imply? -->
The correlation, represented by the symbol $\rho$, measures how much a judge's perception of a specific trait in one stimulus depends on their perception of the same trait in another. As with the discriminal dispersions, this correlation shapes the distribution of the discriminal difference, directly impacting the outcomes of pairwise comparisons. @fig-correlation presents a similar thought experiment as in @sec-theory-issue1a to illustrate this idea. The illustration now assumes that the discriminal dispersions for both texts remain constant. The figure shows that as the correlation between the texts increases, the distribution of their discriminal difference becomes narrower. This narrowing affects the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$, thus influencing the comparison outcome. Furthermore, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less likely to favor Text B over Text A (shaded gray area) compared to scenarios where the texts are highly correlated.

@@

<!-- 2. But this is not the scenario we observe -->
However, in practice, researchers approach this process in reverse. They begin by observing the sample of outcomes favoring Text B over Text A and then use the BTL model to estimate the discriminal difference and the discriminal processes of the stimuli. Given that the BTL model assumes independent discriminal processes across comparisons, if this assumption holds, then the model estimates a discriminal difference distribution that accurately reflects the "true" discriminal difference of the texts. This scenario is illustrated with @fig-correlation when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\rho=0$. Once more, the estimation accuracy of the discriminal difference ensures reliable estimates for the discriminal processes of the texts [(citation needed?)]{style="color:red;"}.

@@

<!-- 2. But does it hold?  -->
Notably, Thurstone attributed the independence of stimuli to the cancellation of potential judges' biases. He argued that this cancellation resulted from two opposing and equally weighted effects occurring during pairwise comparisons [@Thurstone_1927b, pp. 268]. @Andrich_1978 provided a mathematical demonstration of this cancellation using the BTL model under the assumption of discriminal processes with additive biases. However, it is easy to imagine at least two scenarios in which the zero correlation assumption is almost certainly invalid: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli and when an additional hierarchical structure is relevant to the stimuli.

<!-- 3. The first scenario where the assumption does not hold -->
In the first scenario, the intricate aspects of multidimensional, complex traits may introduce dependencies between the stimuli due to certain judges' biases that resist cancellation. Research on text quality suggests that when judges evaluate these traits, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. These additional relevant characteristics, which are unlikely to be equally weighted or opposing, can exert an uneven influence on judges' perceptions, creating biases in their judgments and, ultimately, introducing dependencies between stimuli [@vanderLinden_et_al_2017_II, pp. 346]. For example, this could occur when a judge assessing the argumentative quality of a text places more weight on its grammatical accuracy than other judges, thereby favoring texts with fewer errors but weaker arguments. While direct evidence for this particular scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the notion that the factors influencing pairwise comparisons may not always cancel out.

@@

<!-- 4. The second scenario where the assumption does not hold -->
In the second scenario, the shared context or inherent connections created by additional hierarchical structures may further introduce dependencies between stimuli, a statistical phenomenon commonly known as clustering [@Everitt_et_al_2010]. Despite the CJ literature acknowledging the existence of such hierarchical structures, the statistical handling of this additional source of dependence between stimuli has been inadequate. For instance, when CJ data incorporates multiple samples of stimuli from the same individuals, researchers frequently rely on (average) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. However, this approach can introduce additional statistical and measurement issues, which we discuss in greater detail in @sec-theory-issue2.

<!-- 5. But what are the challenges when this assumption is violated? -->
In any case, similar to @sec-theory-issue1a, model misspecification due to an erroneous assumption of zero correlation between stimuli can lead to significant statistical and measurement issues. For instance, the model may over- or underestimate how accurately the outcome reflects the "true" discriminal differences between stimuli. Such inaccuracies can result in spurious inferences about these differences and, by extension, about the stimuli's discriminal processes. This scenario is illustrated in @fig-correlation, where the model's discriminal difference distribution aligns with the thick continuous line for $\rho=0$, while the "true" discriminal difference follows any discontinuous line where $\rho \neq 0$.

This misspecification may arise from neglecting additional relevant traits, excluding judges based on misfit statistics, or ignoring hierarchical (grouping) structures. Neglecting relevant traits, such as judges' biases, can cause dimensional mismatches in the BTL model, artificially inflating the trait's reliability [@Hoyle_et_al_2023, pp. 341] or, worse, introducing bias into the trait's estimates [@Ackerman_1989]. Excluding judges based on misfit statistics risks discarding valuable information, which may further bias the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. Finally, ignoring hierarchical structures may reduce the precision of model parameter estimates, potentially amplifying the overestimation of the trait's reliability [@Hoyle_et_al_2023, pp. 482].

@@

In aggregate, researchers' inadequate handling of violations to the assumptions of equal dispersion and zero correlation between stimuli, coupled with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, can potentially compromise the reliability of the trait estimates and, by extension, their validity [@Perron_et_al_2015, pp. 2]. Consequently, adopting a more systematic and integrated approach to examine the factors influencing CJ experiments could offer several statistical and measurement benefits, including the ability to address these issues.

@@

This section introduces a theoretical model for Comparative Judgment (CJ) that builds upon Thurstone's theory. The model systematically accounts for the factors influencing judges during CJ experiments. These theoretical and practical insights are then translated into a statistical model.

@@

@@

@@

::: {.panel-tabset group="Figure"}

#### Figure 3

![](/images/png/CJ_TM_A1.png){#fig-CJ_TM_A1 width=20%}

#### Figure 4

![](/images/png/CJ_TM_A2.png){#fig-CJ_TM_A2 width=30%}

#### Figure 5

![](/images/png/CJ_TM_A3.png){#fig-CJ_TM_A3 width=42%}

#### Figure 6

![](/images/png/CJ_TM_A4.png){#fig-CJ_TM_A4 width=42%}

@@

::: panel-tabset

#### Figure 7

![](/images/png/CJ_TM_B1.png){#fig-CJ_TM_B1 width=42%}

#### Figure 8

![](/images/png/CJ_TM_B2.png){#fig-CJ_TM_B2 width=62%}

#### Figure 9

![](/images/png/CJ_TM_B3.png){#fig-CJ_TM_B3 width=72%}

#### Figure 10
